An Algebraic Representation of Wireless Workload

### 2008-02-08 paper.tex
Lot of work...

\newcommand{\thetitle}{Linear Algebra as a Unifying Theory of Workload Modeling}

@@

\newcommand{\FHC}{Hern\'andez-Campos~\textit{et~al.}}
\newcommand{\class}[1]{\textsc{\small{#1}}}
\newcommand{\model}[1]{#1}
\newcommand{\RFC}[1]{\caps{RFC}~{#1}}

@@

\newcommand{\X}{\mathsf{X}}
\newcommand{\M}{\mathsf{M}}
\newcommand{\E}[1]{\left<#1\right>}
\newcommand{\R}{\mathbb{R}}
\newcommand{\Q}{\mathcal{Q}}
\newcommand{\Qt}{Q_{\text{time}}}
\newcommand{\Qi}{Q_{\text{ival}}}
\newcommand{\Di}{Q^{-1}_{\text{ival}}}
\newcommand{\ones}[1]{\mathbf{1}_{#1}}
\newcommand{\zeros}[1]{\mathbf{0}_{#1}}

@@

\newcommand{\TTT}{\text{TTT}}
\newcommand{\TTU}{\text{TTU}}
\newcommand{\TUU}{\text{TUU}}
\newcommand{\TUT}{\text{TUT}}
\newcommand{\UTU}{\text{UTU}}
\newcommand{\UTT}{\text{UTT}}

@@

\begin{abstract}
We present a representation of wireless workload patterns as large, sparse matrices, together with a method for stochastically generating experimental workload from any given matrix. Each row vector represents a single flow of packets in the network, with components representing the \caps{IP} protocol type, source and destination nodes, flow size, start time, and packet behavior. The essential property of the algebraic representation is that the summation of rows naturally yields a faithful description of the aggregate behavior of the corresponding flows. This deceptively simple property allows us to express many common types of traffic models succinctly in terms of a few matrix multiplications and linear constraints. This approach has many benefits: 1) it makes the meaning of certain generally understood but vague concepts, such as ``uniform behavior,'' mathematically precise and unambiguous; 2) it allows us to see clearly, through the lens of linear algebra, the implications of common modeling assumptions; 3) the implementation of traffic models becomes unprecedentedly simple and orthogonal, requiring only a handful of high-level matrix operations, which can be freely composed.
We use the differential simulation methodology introduced in~\cite{Karpinski07:realism} and applied in~\cite{Karpinski07:cbr-failure} to common uniform traffic models, to experimentally validate the general matrix model. 
%We demonstrate experimentally that the general matrix model accurately reproduces network performance metrics, using the differential simulation methodology introduced in~\cite{Karpinski07:realism} and applied in~\cite{Karpinski07:cbr-failure} to common uniform traffic models. A range of common modeling approaches are implemented in the general matrix model and evaluated experimentally.
%
%We demonstrate how linear algebra can be effectively used to represent and perform computations with traffic behaviors in wireless networks. Moreover, we map common concepts and techniques from traffic modeling into this new algebraic domain, showing how they can be expressed with mathematical precision and succinctness, while simultaneously being made clearer and more general. We demonstrate experimentally that the general matrix model accurately reproduces network performance metrics, using the differential simulation methodology introduced in~\cite{Karpinski07:realism} and applied in~\cite{Karpinski07:cbr-failure} to discredit the common constant bit-rate model. A range of common modeling approaches are implemented in the general matrix model and evaluated experimentally. We conclude with future directions for modeling research based on the new algebraic framework.
\end{abstract}

@@

More than twenty years of research in analysis and modeling of network traffic have yielded tremendous advances in our understanding of the complex behaviors that emerge from the interaction of millions of humans and computers on the Internet and in local area networks~(\caps{LAN}s). The rate of packet arrivals at routers on the Internet backbone is known to exhibit complex, self-similar patterns with long-range dependence~\cite{Paxson95,Paxson96,Willinger98}. Other research~\cite{Hernandez06:dissertation} has demonstrated how to accurately model the behavior of a single flow of packets between two application endpoints, including ``closed-loop'' feedback induced by the ubiquitous transmission control protocol~(\caps{TCP}). Recently, the high-level statistical properties of wireless \caps{LAN} traffic have been rigorously modeled with a variety of parametric distributions~\cite{Hernandez06:wlan-traffic}.

Despite great breakthroughs in understanding and methodology, the networking community remains without a ``cookbook'' for generating realistic \caps{LAN} workloads. By ``cookbook,'' we mean an explicit method for generating network traffic---down to the size and send time of each application data unit. In the current standard practice for experimental workload generation, drastic assumptions of uniformity abound. Flows are commonly modeled as constant bit-rate (\caps{CBR}) streams of identical packets sent at an invariant rate; the flows in an experiment are commonly given the same duration and data rate; nodes in the network are often selected as flow endpoints randomly or arbitrarily. While more complex approaches have been proposed~\cite{Sommers04,Avallone04,Avallone06}, % TODO: better references here.
the simplest, most na\"ive models remain standard because no objective criterion exists by which to compellingly argue that any one model is better than the rest. It is simply unclear that simplistic models do any harm, or how, if at all, complex models provide benefit.

We believe that two major breakthroughs are needed to improve traffic modeling. The first requirement is an objective standard of realism by which to compare traffic models. Such a standard has already been introduced in~\cite{Karpinski07:realism} and used to demonstrate that uniform traffic modeling, including the ubiquitous \caps{CBR} model, drastically distort performance metrics at all levels of the network protocol stack~\cite{Karpinski07:cbr-failure}. The second required breakthrough is a unified framework for expressing traffic models that allows researchers to systematically express and understand the assumptions and transformations entailed in each model. The major contribution of this paper is in addressing this second requirement.

%Our previous work on traffic modeling provided the means

In this paper, we present a highly general model for representing detailed full-network traffic patterns algebraically. The General Matrix Model (\caps{GMM}) represents the behavior or each network flow as a vector in a high-dimensional linear space, and represents total network behavior as a large matrix whose rows are these flow-behavior vectors. The crucial property of this representation is that it respects vector space operations: summation of behavior vectors gives an accurate description of the aggregate behavior of the flows represented by those vectors. The benefit of this representation is that it makes it immediately possible to express a broad variety of assumptions and conditions on traffic patterns in terms of simple matrix operations. It also provides vital mathematical clarity and allows us to apply all the tools of modern linear algebra to the problems of traffic modeling.

@@

%It is natural to compare these common, na\"ive models with the high-level parametric models proposed by \FHC~\cite{Hernandez06:wlan-traffic}. Here, however, we face another fundamental challenge in traffic modeling. Even with specific models for various aspects of network behavior, it remains non-trivial to combine them into a coherent composite model capable of generating concrete workloads. The {\FHC} models, for example, specify the distribution of flows per node in the network; since every flow has two end-points, however, it remains unclear how to assign end-points to flows such that the marginal distribution of flow per node matches the BiPareto model indicated. Similarly, the models dictate BiPareto marginal distribution of flow sizes across the entire network, but say nothing about how these sizes are to be distributed among the nodes. We could assume that the number of flows per node is independent from the size of the flows at that node, but, as we shall see, this assumption is invalid. Moreover, the details of flow behavior---packet sizes and inter-packet intervals---remain entirely undefined.

@@

%: objects and their properties together with rules for delegating, inheriting, and overriding the properties across the objects in the \caps{TOM} hierarchy. The \caps{TOM} allows

@@

%Our analysis shows that the standard constant bit-rate (\caps{CBR}) traffic model, with uniformly chosen nodes and uniform flows of traffic, significantly misrepresents important performance metrics at every level of the network. To qualify these results in more depth, we also investigate partially synthetic variants of this standard model, allowing us to reason about which aspects of workload models have the most impact. For the most commonly used model, the delay of application data traversing the network from sender to receiver is underestimated by more than a factor of 5, \textit{on average}, and by more than a factor of 11 in a quarter of usage scenarios. Network control overhead is overestimated, on average, by a factor of about 2.5, while link control overhead is underestimated by almost a factor of 2. Even for metrics where the average and median misrepresentations are not so extreme, the statistical characteristics of error values often indicate that something is fundamentally unrealistic. None of the partially synthetic models manage to accurately represent more than one or two performance characteristics. We further use the \caps{AODV}~\cite{rfc:aodv} and \caps{OLSR}~\cite{rfc:olsr} ad~hoc routing protocols to show that the relative performance of protocols can be inverted when changing from using a simplistic traffic model to using real traffic: using random end-point, uniform \caps{CBR} traffic, \caps{AODV} appears to induce less link-layer overhead than \caps{OLSR}, whereas using real traffic, it in fact induces more.

@@

The rest of the paper is organized as follows. In \Section{motivation} we present motivation and related work. The General Matrix Model is presented in \Section{general-matrix-model}. Our experimental and analytical methodology for evaluating traffic models presented in \Section{methodology}, while the results of our experiments are explained and analyzed in \Section{results}. The ramifications of these results are discussed in \Section{discussion}. Finally, in \Section{conclusions}, we conclude with a discussion of how this research may be applied to current wireless studies, and how it points the way to better traffic models for the future.

@@

The interaction of wireless users and application behavior with the lower layers of the networking stack is characterized by where, when, how much, and to whom data is transmitted. The joint pattern of traffic generation and mobility through time and space completely determines the effect of wireless usage on the lower levels of the network. This is due to the data-agnostic nature of the protocol stack: by design, \caps{IP} networks treat all data in the same manner.%~\cite{Clark88}.
\footnote{This is violated by some quality of service (QoS) schemes. However, we can simply add QoS metadata---such as traffic classes or urgency flags---to our models of user behavior and the rest of our arguments remain valid. The network is still disinterested in the exact content of the data being transported; only the QoS metadata is relevant.} The credibility of conclusions derived from simulation or experimental deployment depends crucially on our confidence that the models used to generate traffic and motion in the experiments are sufficiently realistic.

Paxson and Floyd observed in~\cite{Paxson95} that the interplay between end-point behavior and the network conditions is inherently \textit{closed-loop} in the sense that it is potentially affected by complex feedback. Traffic models typically attempt to preserve the closed-loop behavior of network traffic~\cite{Avallone04,Hernandez06:dissertation}. This presents a fundamental difficulty, however, in that it presumes that we know the intent of end-points: what \textit{would} they have done under different conditions? While we can speculate about what an individual node might hypothetically do, we currently do not understand the impact of the full-network traffic pattern upon performance metrics at all---even without trying to account for hypothetical reactions to alternate situations. Especially in the wireless setting, a fuller understanding of total network behavior must be reached before we can sensibly tackle the complexities of multi-level behavioral feedback. Accordingly, in this paper we attempt to provide a first-order approximation of complete network behavior by studying the response of performance metrics to \textit{open-loop} traffic models without multi-level feedback. It is important to realize that while this does not provide a final picture, we currently lack even a first-order understanding of the effect of different workloads on performance. This first-order understanding is an essential initial step.

@@

\caps{D-ITG}~\cite{Avallone04} generates flows using an independent sampling model for packet sizes and inter-packet intervals. The framework contains pre-made models for several common types of Internet traffic. The focus of this project, however, is on providing the infrastructure to generate very large volumes of synthetic Internet-like backbone traffic. No analysis is provided for determining the realism of traffic mixes, or for choosing flow end-points realistically. In wireless networks, these factors are of crucial importance to performance, and cannot be overlooked. Both Harpoon and \caps{D-ITG} provide excellent traffic generation platforms, but do not provide a systematic framework for understanding or reproducing realistic whole-network workload in the wireless setting.

@@

The traffic workload pattern in a network is a collection of \caps{IP} packets between hosts in the network sent at certain times. It is standard in network analysis, however, to aggregate sequences of packets sharing the same \caps{IP} protocol, source and destination nodes, and source and destination \caps{TCP} or \caps{UDP} port numbers. Such a sequence of packets is called a \textit{flow}. The traffic pattern is characterized completely by the detailed behavior of it's flows. The behavior of a flow is defined by six properties: its source and destination nodes; its start and end times; and the specific sizes of and intervals between its constituent packets. We propose to represent the behavior of each flow in terms of six vectors---one for each aspect of its behavior. The aggregate behavior of all the flows in a workload pattern are then represented as matrices, the rows of which are the behavior vectors of the flows.

@@

The {packets-bytes-duration} representation is clearly overly simplistic and very incomplete. Yet, in giving a description of each flow, it is far more detailed than the most commonly used experimental traffic models. It captures nothing about variability of packet sizes, nor about the ``burstiness'' of packet transmissions; it leaves start and end time unspecified; it says nothing about the source or destination of the flow. At the opposite extreme, it is possible to embed the exact start time of each flow, and the exact size and transmission times of its packets, etc. into a vector space. The difficulty comes in doing so in a manner that makes the summation of vectors correspond naturally and meaningfully to the aggregate behavior of flows. We take a highly detailed but intermediate approach, suggested by the results of~\cite{Karpinski07:realism}: to reproduce network performance accurately, it suffices to know, for each flow, the correct source node, destination node, start time, end time, and empirical distributions of packet sizes and inter-packet intervals. If there are $f$ flows in the entire workload pattern, then the collective behavior can be described with six matrices; the rows of each matrix are the behavior vectors of the flows, indexed in some chosen order. We use the following notation for the behavior matrices and their constituent vectors:\vspace{-0.25em}
\begin{tabbing}
\hspace{1.5em}\=\bullet\hspace{0.5em}\=\hspace{9.75em}\=
$D\:$\=$=\parensx{\vec{d}_i}\:$\=$=\parensx{d_{ij}}\:$\=\kill
\>\bullet\> sources: \>$S$\>$=\parensx{\vec{s}_i}$\>$=\parensx{s_{ij}}$\>$\in\R^{f \times n}$,\\
\>\bullet\> destinations: \>$D$\>$=\parensx{\vec{d}_i}$\>$=\parensx{d_{ij}}$\>$\in\R^{f \times n}$,\\
\>\bullet\> start times: \>$B$\>$=\parensx{\vec{b}_i}$\>$=\parensx{b_{ij}}$\>$\in\R^{f \times d_t}$,\\
\>\bullet\> end times: \>$E$\>$=\parensx{\vec{e}_i}$\>$=\parensx{e_{ij}}$\>$\in\R^{f \times d_t}$,\\
\>\bullet\> packet sizes: \>$Z$\>$=\parensx{\vec{z}_i}$\>$=\parensx{z_{ij}}$\>$\in\R^{f \times d_z}$,\\
\>\bullet\> inter-packet intervals: \>$V$\>$=\parensx{\vec{v}_i}$\>$=\parensx{v_{ij}}$\>$\in\R^{f \times d_v}$.
\end{tabbing}
The total behavior is the direct product of these components: $F = \seq{S,D,B,E,Z,V} \in \R^{f \times \Sigma}$ with $\Sigma=2n+2d_t+d_z+d_v$. Many common traffic models can be expressed succinctly and precisely in terms of transformations or conditions on this behavior matrix and its constituents. In the following sections, we explain in detail how each flow's behavior is represented in these matrices. Finally, we describe how a model instance stochastically generates network traffic.

@@

Let $n$ be the number of nodes in the traffic collection to be modeled. We let $s_{ij}=1$ if the $j$th node is the source of the $i$th flow, and zero otherwise; similarly, $d_{ik}=1$ if the destination of the $i$th flow is the $k$th node. This sparse matrix may seem to be an overly expansive representation, given that start and end nodes can be represented simply as indices from $\set{1,\dots,n}$. The advantage becomes clear when we consider vector operations aggregating collections of behavior vectors. The sum of the source indices over a collection is meaningless; the sum of a set of source vectors, $\set{\vec{s}_i}$, however, is an $n$-dimensional histogram indicating how often each source node occurs in the collection.

\subsection{Start \& End Times}

We represent start and end times using indicator vectors. Unlike node indices, however, time is continuous, so we must first map times onto a range of time indices $\set{1,\dots,d_t}$. We transform the range $[0,t_{\max}]$ of possible start times using a simple quantization function:
\begin{align}
\label{eqn:time-quantization}
\Qt(t) &= \max\set{1,\ceil{d_t\fracpx{t}{t_{\max}}}}.
\end{align}
%Suppose the $i$th flow has a quantized start index of $j$ and a end index of $k$. Then $b_{ij'}=\delta{jj'}$ and $e_{ik'}=\delta_{kk'}$.
Let $j$ and $k$ be the quantized start and end indices of a flow. Then the start and end vectors for the flow are $\vec{b}=\mathbf{e}_j$ and $\vec{e}=\mathbf{e}_k$ ($\mathbf{e}_j$ has 1 at index $j$ and zeros elsewhere). Again, this may seem excessively complex, but the advantage becomes clear when we aggregate flows: the sum of start vectors over a collection of flows is a $d_t$-dimensional histogram indicating the frequency of start times occurring in each quantization bin.

>>>>

\newcommand{\thetitle}{An Algebraic Representation of Wireless Workload}

@@

\newcommand{\FHC}{Hern\'andez-Campos~\textit{et~al.}}
\newcommand{\class}[1]{\textsc{\small{#1}}}
\newcommand{\model}[1]{\textsf{\small{#1}}}
\newcommand{\RFC}[1]{\caps{RFC}~{#1}}

@@

\newcommand{\X}{\mathsf{X}}
\newcommand{\M}{\mathsf{M}}
\newcommand{\E}[1]{\left<#1\right>}
\newcommand{\R}{\mathbb{R}}
\newcommand{\Q}{\mathcal{Q}}
\newcommand{\Qt}{Q_{\text{time}}}
\newcommand{\Qb}{Q_{\text{byte}}}
\newcommand{\Qi}{Q_{\text{ival}}}
\newcommand{\Di}{Q^{-1}_{\text{ival}}}
\newcommand{\ones}[1]{\mathbf{1}_{#1}}
\newcommand{\zeros}[1]{\mathbf{0}_{#1}}

@@

@@

\begin{abstract}
We propose a representation of wireless workload patterns as large, sparse matrices and provide a method for stochastically generating experimental workload from a given matrix. The essential property of the algebraic representation is that the summation of vectors naturally yields a faithful description of the aggregate behavior of the corresponding flows. This deceptively simple property allows us to express many common concepts from traffic modeling succinctly in terms of a few linear transformations. The algebraic representation has many benefits: 1)~it makes the meaning of generally understood but vague concepts, such as ``uniform behavior,'' mathematically precise and unambiguous; 2)~it allows us to see clearly, through the lens of linear algebra, the implications of common modeling assumptions; 3)~the implementation of traffic models becomes unprecedentedly simple and orthogonal, requiring only a handful of high-level matrix operations, which can be freely composed; 4)~the vast body of algebraic theory and highly optimized numerical linear software may immediately be applied to traffic modeling. We use the paired differential simulation methodology introduced by the authors in previous work to experimentally demonstrate that the general matrix model accurately reproduces realistic network performance~\cite{Karpinski07:realism,Karpinski07:cbr-failure}. We use the same experimental methodology to explore the implications of various assumptions and simplifications that are commonly made in traffic modeling.
%We demonstrate experimentally that the general matrix model accurately reproduces network performance metrics, using the differential simulation methodology introduced in~\cite{Karpinski07:realism} and applied in~\cite{Karpinski07:cbr-failure} to common uniform traffic models. A range of common modeling approaches are implemented in the general matrix model and evaluated experimentally.
%
%We demonstrate how linear algebra can be effectively used to represent and perform computations with traffic behaviors in wireless networks. Moreover, we map common concepts and techniques from traffic modeling into this new algebraic domain, showing how they can be expressed with mathematical precision and succinctness, while simultaneously being made clearer and more general. We demonstrate experimentally that the general matrix model accurately reproduces network performance metrics, using the differential simulation methodology introduced in~\cite{Karpinski07:realism} and applied in~\cite{Karpinski07:cbr-failure} to discredit the common constant bit-rate model. A range of common modeling approaches are implemented in the general matrix model and evaluated experimentally. We conclude with future directions for modeling research based on the new algebraic framework.
\end{abstract}

@@

More than twenty years of research in analysis and modeling of network traffic have yielded tremendous advances in our understanding of the complex behaviors that emerge from the interaction of millions of humans and computers on the Internet and in local area networks~(\caps{LAN}s). The rate of packet arrivals at routers on the Internet backbone is known to exhibit complex, self-similar patterns with long-range dependence~\cite{Paxson95,Paxson96,Willinger98}. Other research has demonstrated how to accurately model the behavior of a single flow of packets between two application endpoints, including ``closed-loop'' feedback induced by the ubiquitous transmission control protocol~(\caps{TCP})~\cite{Hernandez06:dissertation}. Recently, the high-level statistical properties of wireless \caps{LAN} traffic have been rigorously modeled with a variety of parametric distributions~\cite{Hernandez06:wlan-traffic}.

Despite great breakthroughs in understanding and methodology, the networking community remains without a ``cookbook'' for generating realistic \caps{LAN} workloads. By ``cookbook,'' we mean an explicit method for generating network traffic---down to the size and transmission time of each application data unit. This lack is particularly relevant to wireless research, since the medium fundamentally resource-limited, and consequently local traffic patterns can have a huge impact on overall network performance~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.

In the current standard practice for experimental workload generation, drastic assumptions of uniformity abound. Flows are commonly modeled as constant bit-rate (\caps{CBR}) streams of identical packets sent at an invariant rate. The flows in an experiment are commonly given the same duration and data rate. Nodes in the network are often selected as flow endpoints randomly or arbitrarily. While more complex approaches have been proposed~\cite{Sommers04,Avallone04,Avallone06}, % TODO: better references here.
the simplest, most na\"ive models remain standard because no objective criterion exists by which to compellingly argue that any one model is better than the rest. It is simply unclear that simplistic models do any harm, or how, if at all, complex models provide benefit.

We believe that two major breakthroughs are needed to improve traffic modeling. The first requirement is an objective standard of realism by which to compare traffic models. Such a standard has already been introduced and used to demonstrate that uniform traffic modeling, including the ubiquitous \caps{CBR} model, drastically distort performance metrics at all levels of the network protocol stack~\cite{Karpinski07:realism,Karpinski07:cbr-failure}. This standard serves to demonstrate the harm of overly simplistic models, such as \caps{CBR}, and provides the means to test whether more complex models deliver the desired improvements.
The second breakthrough that is needed to improve modeling is a unified framework for expressing a broad variety of traffic models, that allows researchers to systematically express and reason about the assumptions and simplifications entailed in each model.

The major contribution of this paper is addressing this second requirement. We propose the General Matrix Model (\caps{GMM}) for local network traffic.  In this model, the traffic pattern of an entire network is represented as a large, sparse matrix. Each row vector represents a single flow of packets in the network, with components representing the \caps{IP} protocol type, source and destination nodes, flow size, start time, and packet behavior.  The crucial property of this representation is that it respects vector space operations: summation of behavior vectors gives an accurate description of the aggregate behavior of the flows represented by those vectors. One benefit of this representation is that it makes it immediately possible to express a broad variety of assumptions and conditions on traffic patterns in terms of simple matrix operations. It also provides vital mathematical clarity and allows us to apply the theoretical and computational tools of modern linear algebra to the problems of traffic analysis and modeling.

After presenting \caps{GMM} and discussing how common concepts in traffic modeling can be expressed in precise mathematical terms within this framework, we proceed with experimental validation of \caps{GMM}. We use the technique of \textit{paired differential simulation} that the authors proposed and used in previous work~\cite{Karpinski07:realism,Karpinski07:cbr-failure}. This technique uses real wireless trace traffic as an experimental control and conducts paired experiments comparing network performance from simulations using control traffic against others using synthetically generated traffic. The results are positive: \caps{GMM} accurately reproduces network performance characteristics at all levels of the protocol stack.
%This does not solve the ``cookbook'' problem, however, because the \caps{GMM} must still be based on real traffic in order to reproduce its behavioral characteristics. It does however, reduce the problem of general traffic modeling to a more constrained problem of matrix modeling.

Our experiments with various common assumptions of uniformity and regularity in network behavior yield mixed conclusions: most straightforward assumptions of homogenous behavior distort performance results significantly, but there are notable exceptions. The clearest exception is the time domain: traffic models appear to be relatively insensitive to local perturbation of the start time of flows---so long as the overall distribution remains approximately correct. Our conclusion is that traffic modeling needs to discard its assumptions of uniformity, regularity and independence. Instead we need to move ahead trying to understand the complex dependencies between the many aspects of traffic behaviors. By reducing traffic modeling to the well-understood domain of numerical linear analysis, the General Matrix Model brings us significantly closer to the ultimate goal of having a ``cookbook'' model for generating arbitrary, realistic experimental workload patterns.

@@

%It is natural to compare these common, na\"ive models with the high-level parametric models proposed by \FHC~\cite{Hernandez06:wlan-traffic}. Here, however, we face another fundamental challenge in traffic modeling. Even with specific models for various aspects of network behavior, it remains non-trivial to combine them into a coherent composite model capable of generating concrete workloads. The {\FHC} models, for example, specify the distribution of flows per node in the network; since every flow has two endpoints, however, it remains unclear how to assign endpoints to flows such that the marginal distribution of flow per node matches the BiPareto model indicated. Similarly, the models dictate BiPareto marginal distribution of flow sizes across the entire network, but say nothing about how these sizes are to be distributed among the nodes. We could assume that the number of flows per node is independent from the size of the flows at that node, but, as we shall see, this assumption is invalid. Moreover, the details of flow behavior---packet sizes and inter-packet intervals---remain entirely undefined.

@@

% objects and their properties together with rules for delegating, inheriting, and overriding the properties across the objects in the \caps{TOM} hierarchy. The \caps{TOM} allows

@@

%Our analysis shows that the standard constant bit-rate (\caps{CBR}) traffic model, with uniformly chosen nodes and uniform flows of traffic, significantly misrepresents important performance metrics at every level of the network. To qualify these results in more depth, we also investigate partially synthetic variants of this standard model, allowing us to reason about which aspects of workload models have the most impact. For the most commonly used model, the delay of application data traversing the network from sender to receiver is underestimated by more than a factor of 5, \textit{on average}, and by more than a factor of 11 in a quarter of usage scenarios. Network control overhead is overestimated, on average, by a factor of about 2.5, while link control overhead is underestimated by almost a factor of 2. Even for metrics where the average and median misrepresentations are not so extreme, the statistical characteristics of error values often indicate that something is fundamentally unrealistic. None of the partially synthetic models manage to accurately represent more than one or two performance characteristics. We further use the \caps{AODV}~\cite{rfc:aodv} and \caps{OLSR}~\cite{rfc:olsr} ad~hoc routing protocols to show that the relative performance of protocols can be inverted when changing from using a simplistic traffic model to using real traffic: using random endpoint, uniform \caps{CBR} traffic, \caps{AODV} appears to induce less link-layer overhead than \caps{OLSR}, whereas using real traffic, it in fact induces more.

@@

The rest of the paper is organized as follows. In \Section{motivation} we discuss motivation and related work. The General Matrix Model is presented in \Section{general-matrix-model}. Our experimental and analytical methodology for evaluating traffic models is presented in \Section{methodology}, while the results of our experiments are explained and analyzed in \Section{results}. The ramifications of these results are discussed in \Section{discussion}. Finally, in \Section{conclusions}, we conclude with a discussion of how this research may be applied to current wireless studies, and how it points the way to better traffic models for the future.

@@

The interaction of wireless users and application behavior with the lower layers of the networking stack is characterized by where, when, how much, and to whom data is transmitted. The joint pattern of traffic generation and mobility through time and space completely determines the effect of wireless usage on the lower levels of the network. This is due to the data-agnostic nature of the protocol stack: by design, \caps{IP} networks treat all data in the same manner.%~\cite{Clark88}.
\footnote{This is violated by some quality of service (QoS) schemes. However, we can simply add QoS metadata---such as traffic classes or urgency flags---to our models of user behavior and the rest of our arguments remain valid. The network is still disinterested in the exact content of the data being transported; only the QoS metadata is relevant.} The credibility of conclusions derived from simulation or experimental deployment depends crucially on our confidence that the models used to generate traffic in experiments are sufficiently realistic.

Paxson and Floyd observed in~\cite{Paxson95} that the interplay between endpoint behavior and the network conditions is inherently \textit{closed-loop} in the sense that it is potentially affected by complex feedback. Traffic models typically attempt to preserve the closed-loop behavior of network traffic~\cite{Avallone04,Hernandez06:dissertation}. This presents a fundamental difficulty, however, in that it presumes that we know the intent of endpoints: what \textit{would} they have done under different conditions? While we can speculate about what an individual node might hypothetically do, we currently hardly understand the impact of the full-network traffic pattern upon performance metrics at all---even without trying to account for hypothetical reactions to alternate situations. Especially in the wireless setting, a fuller understanding of total network behavior must be reached before we can sensibly tackle the complexities of multi-level behavioral feedback. Accordingly, in this paper we attempt to provide a first-order approximation of complete network behavior by studying the response of performance metrics to \textit{open-loop} traffic models without multi-level feedback. It is important to realize that while this does not provide a final picture, we currently lack even a first-order understanding of the effect of different workloads on performance. This first-order understanding is an essential initial step.

@@

\caps{D-ITG}~\cite{Avallone04} generates flows using an independent sampling model for packet sizes and inter-packet intervals. The framework contains pre-made models for several common types of Internet traffic. The focus of this project, however, is on providing the infrastructure to generate very large volumes of synthetic Internet-like backbone traffic. No analysis is provided for determining the realism of traffic mixes, or for choosing flow endpoints realistically. In wireless networks, these factors are of crucial importance to performance, and cannot be overlooked. Both Harpoon and \caps{D-ITG} provide excellent traffic generation platforms, but do not provide a systematic framework for understanding or reproducing realistic whole-network workload in the wireless setting.

@@

The traffic workload pattern in a network is a collection of \caps{IP} packets between hosts in the network sent at certain times. It is standard in network analysis, however, to aggregate sequences of packets sharing the same \caps{IP} protocol, source and destination nodes, and source and destination \caps{TCP} or \caps{UDP} port numbers. Such a sequence of packets is called a \textit{flow}. The traffic pattern of an entire network is simply the collected behaviors of all flows occurring in the network. The behavior of a flow as it affects the network is characterized by the following properties: its \caps{IP} protocol type (\caps{TCP}, \caps{UDP}, \caps{ICMP}, etc.); its source and destination nodes; its start time; and the specific sizes of and intervals between its constituent packets. We propose to represent the behavior of each flow in terms of seven vectors, which suffice to describe all of the above properties in enough detail to stochastically generate realistic traffic. The aggregate behavior of all the flows in a workload pattern are then represented as matrices, the rows of which are the behavior vectors of the flows.

@@

The {packets-bytes-duration} representation is clearly overly simplistic and very incomplete. It captures nothing about variability of packet sizes, nor about the ``burstiness'' of packet transmissions; it leaves start time and flow size unspecified; it says nothing about the source or destination of the flow. At the opposite extreme, it is possible to embed the exact start time of each flow, and the exact size and transmission times of its packets, etc. into a vector space. The difficulty comes in doing so in a manner that makes the summation of vectors correspond naturally and meaningfully to the aggregate behavior of flows. We take a highly detailed but intermediate approach, suggested by the results of~\cite{Karpinski07:realism}: to reproduce network performance accurately, it suffices to know, for each flow, the flow type, source node, destination node, start time, flow size, and empirical distributions of packet sizes and inter-packet intervals. If there are $f$ flows in the entire workload pattern, then the collective behavior can be described with six matrices; the rows of each matrix are the behavior vectors of the flows, indexed in some chosen order. We use the following notation for the behavior matrices and their constituent vectors:\vspace{-0.25em}
\begin{tabbing}
\hspace{1.5em}\=\bullet\hspace{0.5em}\=\hspace{9.75em}\=
$D\:$\=$=\parensx{\vec{d}_i}\:$\=$=\parensx{d_{ij}}\:$\=\kill
\>\bullet\> type: \>$T$\>$=\parensx{\vec{t}_i}$\>$=\parensx{t_{ij}}$\>$\in\R^{f \times 255}$,\\
\>\bullet\> sources: \>$S$\>$=\parensx{\vec{s}_i}$\>$=\parensx{s_{ij}}$\>$\in\R^{f \times n}$,\\
\>\bullet\> destinations: \>$D$\>$=\parensx{\vec{d}_i}$\>$=\parensx{d_{ij}}$\>$\in\R^{f \times n}$,\\
\>\bullet\> start times: \>$A$\>$=\parensx{\vec{a}_i}$\>$=\parensx{a_{ij}}$\>$\in\R^{f \times d_t}$,\\
\>\bullet\> sizes in bytes: \>$B$\>$=\parensx{\vec{b}_i}$\>$=\parensx{b_{ij}}$\>$\in\R^{f \times d_b}$,\\
\>\bullet\> packet sizes: \>$Z$\>$=\parensx{\vec{z}_i}$\>$=\parensx{z_{ij}}$\>$\in\R^{f \times d_z}$,\\
\>\bullet\> inter-packet intervals: \>$V$\>$=\parensx{\vec{v}_i}$\>$=\parensx{v_{ij}}$\>$\in\R^{f \times d_v}$.
\end{tabbing}
The total behavior is the direct product of these components: $F = \seq{T,S,D,A,B,Z,V} \in \R^{f \times \Sigma}$ where $\Sigma=256+2n+d_t+d_b+d_z+d_v$. Many common traffic models can be expressed succinctly and precisely in terms of transformations or conditions on this behavior matrix and its constituents. In the following sections, we explain in detail how each flow's behavior is represented in these matrices. Finally, we describe how a model instance stochastically generates network traffic.

\subsection{Flow Type}

The type of a flow is its \caps{IP} protocol number. This ranges in possible values from 1 to 255 (0 is reserved for \caps{IPv6}). We represent the flow type as an indicator vector: a 255-dimensional vector consisting of all zeros except for the index which is the flow type. So a \caps{TCP} flow, having protocol type 6, would have the type vector $\vec{t}=\seq{0,0,0,0,0,1,0,\dots,0}$. This representation may immediately strike one as unnecessarily expansive and inefficient since a single byte suffices to represent the protocol type. The necessity of this representation arises when we represent aggregations of flows as sums of vectors. The sum of protocol numbers is meaningless: two \caps{TCP} flows plus five \caps{ICMP} flows does not equal a single \caps{UDP} flow, despite the fact that $2\cdot6+5\cdot1=17$. The summation of the corresponding indicator vectors, however, is $\seq{5,0,0,0,0,2,0,\dots}$. This vector is a histogram indicating the frequency with which each protocol type occurred in the collection of flows over which the sum was taken.

@@

Let $n$ be the number of nodes in the traffic collection to be modeled. We let $s_{ij}=1$ if the $j$th node is the source of the $i$th flow, and zero otherwise; similarly, $d_{ik}=1$ if the destination of the $i$th flow is the $k$th node. This sparse matrix may again seem to be an overly expansive representation, given that start and end nodes can be represented simply as indices from $\set{1,\dots,n}$. The sum of a set of source vectors, $\set{\vec{s}_i}$, however, is an $n$-dimensional histogram indicating how often each source node occurs in the collection.

\subsection{Start Time}

We represent start time of each flow using an indicator vector. Unlike node indices, however, time is continuous, so we must first map times onto a range of time indices $\set{1,\dots,d_t}$. We transform the range $[0,t_{\max}]$ of possible start times using a simple quantization function:
\begin{align}
\label{eqn:time-quantization}
\Qt(t) &= \max\set{1,\ceil{d_t\fracpx{t}{t_{\max}}}}.
\end{align}
%Suppose the $i$th flow has a quantized start index of $j$ and a end index of $k$. Then $b_{ij'}=\delta{jj'}$ and $e_{ik'}=\delta_{kk'}$.
Let $k$ be the quantized start index of a flow. Then the start vectors for the flow is $\vec{a}=\mathbf{e}_k$ ($\mathbf{e}_k$ has 1 at index $k$ and zeros elsewhere). The advantage of this representation becomes clear when we aggregate flows: the sum of start vectors over a collection of flows is a $d_t$-dimensional histogram indicating the frequency of start times occurring in each quantization bin. In practice, we use $d_t=1500$.

\subsection{Flow Size}
